---
title: "Oslo Sports Trauma Research Centre R package"
author:
  - Lena Kristin Bache-Mathiesen^[Oslo Sports Trauma Research Centre, lenakba22@gmail.com]
  - Benjamin Clarsen^[Oslo Sports Trauma Research Centre, benjaminc@nih.no]
output: github_document
---
```{r packages, warning=FALSE, echo = FALSE, message=FALSE}
library(tidyverse) # dplyr, tidyr, ggplot2, purrr and everything else you need to survive
library(ostrc)
```

## Overview
The OSTRC package is a collection of R functions 
for working with sports injury data. Among other 
functions, the package provides tools for handling data from the 
Oslo Sports Trauma Research Centre 
Questionnaire on overuse injuries or health problems. 

### OSTRC questionnaire
The OSTRC questionnaire collects self-reported injury and illness data.   
The functions for handling OSTRC-questionnaire data are 
generalized to work with any OSTRC-questionnaire type and version.
The overuse questionnaire is denoted OSTRC-O. 
The health problem questionnaire is denoted OSTRC-H. 
Version 1.0 corresponds to the original introduction of the questionnaires, 
version 2.0 to the updated version after the review panel meeting.
Based on the articles they were introduced, 
in chronological order from the earliest to latest version:

<ol>
<li>Original overuse questionnaire: 
[OSTRC-O 1.0](doi.org/10.1136/bjsports-2012-091524)</li> 
<li>The expanded version capturing all health problems:
[OSTRC-H 1.0](doi.org/10.1136/bjsports-2012-092087)</li>
<li>The review panel meeting updating both questionnaires:
[OSTRC-O 2.0 and OSTRC-H 2.0](doi.org/10.1136/bjsports-2019-101337)</li>
</ol>

## Installation

```{r install_guide, echo = TRUE, warning = FALSE, eval = FALSE}
# Requires the devtools package
# If devtools is not installed, run code:
# install.packages("devtools") before installing ostrc
devtools::install_github("lenakba/ostrc")
```

## Functions
Below is a brief tutorial on helpful functions in the R package. 

### OSTRC questionnaire functions

#### Find health problems and substantial health problems

Given a vector of OSTRC responses to question 1, 
The function `find_hp` identifies whether or not an observation
is a health problem as per the definition of a health problem
in the OSTRC questionnaire. 

```{r find_hp_example, echo = TRUE, warning = FALSE}
ostrc_1 = c(0, 8, 17, 25)

find_hp(ostrc_1)
```

If the vector of responses to question 1 is not
coded with the standard 0, 8, 17, 25 responses,
but with 1 to 4, or 0 to 3, these can also work.
In these cases, `find_hp` uses the `standardize_codes` 
function automatically.

```{r find_hp_example_codes, echo = TRUE, warning = FALSE}
ostrc_1_other_values = c(0, 1, 2, 3)

find_hp(ostrc_1_other_values)
```

Substantial health problems are more 
complicated to identify, and depends on
the OSTRC version (1.0 or 2.0). 
The function `find_hp_substantial`
identifies substantial injuries given 
question 1, 2, and 3 of the OSTRC questionnaire.
If there is no health problem, it will return as missing.

```{r find_hp_substantial_example, echo = TRUE, warning = FALSE}
ostrc_1 = c(0, 8, 8, 8)
ostrc_2 = c(0, 0, 0, 25)
ostrc_3 = c(0, 0, 17, 0)
   
find_hp_substantial(ostrc_1, ostrc_2, ostrc_3)
```

Make note of how `find_hp_substantial` treats missing in one
or more of the OSTRC-questionnaire items.
If one of the three questions indicate a substantial health problem,
`find_hp_substantial` will consider it substantial, 
regardless of missing values
in other items. No matter whether the missing values are, in reality,
substantial or non-substantial, we still have enough information
to conclude that the observation is substantial.

However, if the only non-missing responses
are values that do not indicate a substantial health problem,
then it will return as missing. In these cases, there is no way of knowing
whether one of the missing observations was, in reality,
a response pertaining to a substantial health problem.

```{r find_hp_substantial_example_missing, echo = TRUE, warning = FALSE}
ostrc_1_missing = c(25, NA, NA, NA)
ostrc_2_missing = c(NA, NA, NA, 17)
ostrc_3_missing = c(NA, 8, NA, NA)

find_hp_substantial(ostrc_1_missing, ostrc_2_missing, ostrc_3_missing)
```

#### Calculate severity score and timeloss

Severity scores can also be calculated with `calc_severity_score`.

```{r calc_severity_score_example, echo = TRUE, warning = FALSE}
ostrc_1 = c(0, 8, 8, 8)
ostrc_2 = c(0, 0, 0, 25)
ostrc_3 = c(0, 0, 17, 0)
ostrc_4 = c(0, 0, 25, 25)
   
calc_severity_score(ostrc_1, ostrc_2, ostrc_3, ostrc_4)
```

Finally, one can determine the timeloss of each health problem 
with `calc_timeloss`. 
You receive a vector with the number of weeks lost 
per health problem case in the data. In the example below,
there are 3 case IDs representing 3 health problem cases,
and we receive 3 values representing the number of weeks lost.

```{r calc_timeloss_example, echo = TRUE, warning = FALSE}
d_ostrc = tribble(~id_participant, ~id_case, ~date_sent, ~q1,
                  1, 1, "2023-01-01", 8,
                  1, 1, "2023-01-07", 8,
                  1, 1, "2023-01-14", 8,
                  1, 1, "2023-01-21", 25,
                  1, 18, "2022-12-07", 25,
                  1, 18, "2022-12-14", 25,
                  2, 2, "2023-01-12", 8,
                  3, NA, "2022-06-05", 0)
   
calc_timeloss(d_ostrc, id_participant, id_case, date_sent, q1)
```

#### Create case data

The function `create_case_data` is a helper function that brings 
all the information you need on each health problem in one dataset. 
It finds health problems, substantial health problems, 
severity score, startdate, enddate, and duration (in weeks).
Returns a dataframe where one row describes one unique health problem.
Any other information on the health problem (type, location etc.) is retained. 

```{r create_case_data_example, echo = TRUE, warning = FALSE}
library(tidyverse) # for tribble() and pipe %>% 
d_ostrc = tribble(~id_participant, ~id_case, ~date_ostrc, ~q1, ~q2, ~q3, ~q4, ~hb_type, ~inj_type,
                  1, 1, "2023-01-01", 8, 0, 17, 25, "Injury", "Overuse",
                  1, 1, "2023-01-07", 8, 0, 17, 25, "Injury", "Overuse",
                  1, 1, "2023-01-14", 8, 0, 17, 0, "Injury", "Overuse",
                  1, 18, "2022-12-07", 25, 0, 0, 0, "Illness", NA,
                  2, 2, "2023-01-12", 8, 8, 0, 0, NA, NA,
                  3, 3, "2022-06-05", 0, 0, 0, 0, NA, NA,
                  4, 4, "2023-01-01", 8, 8, 8, 0, "Injury", "Acute")

# note that the date column must be of class date
d_ostrc = d_ostrc %>% mutate(date_ostrc = as.Date(date_ostrc))

# functions returns one row per health problem
# if a response is not a health problem (like participant ID 3, case ID 3)
# it will not be included in the returned data frame.
d_cases = create_case_data(d_ostrc, id_participant, id_case, date_ostrc, q1, q2, q3, q4)
d_cases
```

Any extra columns in the dataset will be included at the end, 
like columns `hb_type` and `inj_typ` in the example above. Below,
we show how these columns appear after being handled by
the `create_case_data` function.

```{r show_all_cols, echo = TRUE, warning = FALSE}
d_cases %>% select(id_participant, id_case, date_ostrc, hb_type, inj_type)
```

### Functions for handling injury data in general

#### Calculate prevalence

The OSTRC package has three functions for calculating the prevalence. 
`calc_prevalence` calculates the prevalence per given timepoint. 
The timepoint can be given by dates, or numeric values, 
such as study week number. The example in this readme is 
an OSTRC-questionnaire sent weekly. `calc_prevalence_mean` calculates the mean prevalence, 
after calculating the weekly prevalence with `calc_prevalence`. 
Meaning in the exmaple below, we will obtain the mean weekly prevalence.
Finally, `calc_prevalence_all` calculates the prevalence of each given health problem type. 
This is practical in cases where you have multiple types of health problems, 
such as illnesses, injuries, contact injuries, acute injuries etc. 
and you wish to calculate the prevalence for each type.

```{r calc_prevalence_example_data, echo = TRUE, warning = FALSE}
# Here we have some example data
# note that we assume the date the questionnaire was sent
# was the day that respondents replied
d_injuries = tribble(~id_participant, ~date_sent, ~injury, ~injury_substantial, ~gender,
                  1, "2023-01-07", 1, 0, "Male",
                  1, "2023-01-14", 1, 1, "Male",
                  1, "2023-01-21", 0, 0, "Male",
                  2, "2023-01-07", 1, 1, "Female",
                  2, "2023-01-14", 0, 1, "Female",
                  2, "2023-01-21", 1, 0, "Female",
                  3, "2023-01-07", 1, 0, "Male",
                  3, "2023-01-14", 0, 0, "Male",
                  4, "2023-01-07", 0, 0, "Male",
                  4, "2023-01-14", 1, 0, "Male",
                  4, "2023-01-21", 1, 0, "Male",
                  5, "2023-01-07", 1, 0, "Female",
                  5, "2023-01-14", 1, 0, "Female",
                  6, "2023-01-07", 1, 1, "Female",
                  6, "2023-01-14", 1, 1, "Female",
                  6, "2023-01-21", 1, 0, "Female",
                  )
```

Below, we calculate the weekly injury prevalence on the example data, 
with the help of `calc_prevalence`. We need to provide the dataframe, 
the column with participant or player ID, 
the date or time the OSTRC questionnaire was sent, 
and the health problem column.

```{r calc_prevalence_example, echo = TRUE, warning = FALSE}
calc_prevalence(d_injuries, id_participant, date_sent, injury)
```

However, sometimes we are only interested in the mean weekly prevalence. 
`calc_prevalence_mean` provides the mean, standard deviation, and confidence intervals. 
You can choose confidence interval level, and the default is at 95%. 
Note that the function arguments are the same as for `calc_prevalence`. 

```{r calc_prevalence_mean_example, echo = TRUE, warning = FALSE}
calc_prevalence_mean(d_injuries, id_participant, date_sent, injury)
```

When you have multiple health problem classifications, it can be
tedious to calculate the prevalence manually for each one. With 
`calc_prevalence_all` you can do them all in one go, with one line of code.
The name of each health problem column needs to be provided as a vector of strings, 
like in the example below.

```{r calc_prevalence_all_example, echo = TRUE, warning = FALSE}
calc_prevalence_all(d_injuries, id_participant, date_sent, c("injury", "injury_substantial"))
```

In some cases, you may wish to calculate prevalences for each category in a group. 
For instance, for males and females separately. Below is an example.

```{r calc_prevalence_all_group_example, echo = TRUE, warning = FALSE}
calc_prevalence_all(d_injuries, id_participant, date_sent, c("injury", "injury_substantial"), "gender")
```

#### Calculate incidence

Similar to prevalence, calculating incidence also has three functions:
`calc_incidence`, `calc_incidence_mean` and `calc_incidence_all`. 
The functions check whether or not the participant had
a health problem of the given type in the previous response. If they did,
any response of a continued health problem will not be considered a new case.
If they did not, a response of a health problem will be considered a new health problem,
and enter the numerator of the incidence calculation. 
Weekly incidences can therefore only be calculated if there are no gaps in the data, 
i.e. if the questionnaire was sent bi-weekly. If the questionnaire was sent monthly, 
this will provide a rough estimate of the monthly incidence.

Using the same example as the prevalence calculations, here is 
the weekly incidence:

```{r calc_incidence_example, echo = TRUE, warning = FALSE}
calc_incidence(d_injuries, id_participant, date_sent, injury)
```

Note that the first row of incidence is missing data. 
This is because we do not have 
health problem information about any of the individuals before
the study started. We therefore cannot know if a health problem response
is a new or continued health problem. The only exception
is if all of the responses are 0 for no health problem,
then we can know for sure that the incidence is 0%. Below is an example.

```{r calc_incidence_example_0start, echo = TRUE, warning = FALSE}
 d_0atstart = tribble(~id_participant, ~week_nr, ~hp,
                              1, 1, 0,
                              1, 2, 0,
                              1, 3, 1,
                              2, 1, 0,
                              2, 2, 0,
                              2, 3, 1)
calc_incidence(d_0atstart, id_participant, week_nr, hp)
```

As with prevalence, we can calculate the mean incidence.

```{r calc_incidence_mean_example, echo = TRUE, warning = FALSE}
calc_incidence_mean(d_injuries, id_participant, date_sent, injury)
```

And obtain the mean incidence for each health problem type, for a group.

```{r calc_eincidence_all_group_example, echo = TRUE, warning = FALSE}
calc_incidence_all(d_injuries, id_participant, date_sent, c("injury", "injury_substantial"), "gender")
```

#### Find and add event IDs

Function `add_event_id` finds intervals for each event in a longitudinal dataset, and
adds a column indexing them. The intervals may be of unequal lengths. 
This is useful for finding intervals between multiple injuries or
episodes of pain on the same individual, that is, identifying injury cases. 
Or the opposite: the intervals where the participant is currently injured or in pain.

It can also be used to find the
so-called micro-cycle between matches and competition events. 
Micro-cycles are the time from a previous competition or match
to the next competition or match. That is,
all days up to first match/competition in one cycle,
all days from first match to second match in the next cycle and so on. 
Coaches often plan training according to such 
micro-cycles.

This is handy for counting the number of injury cases or number of micro-cycles,
calculating the average interval time, 
or calculating the average of anything else per injury or other event.   

```{r event_id_example, echo = TRUE, warning = FALSE}
library(tidyverse)
library(ostrc)
d_matches_sim = tribble(~date, ~match,
                          "2017-08-22", 0,
                          "2017-08-23", 1,
                          "2017-08-24", 0,
                          "2017-08-25", 0,
                          "2017-08-26", 1)
  d_matches_sim = d_matches_sim %>% mutate(date = lubridate::as_date(date))
  add_event_id(d_matches_sim, date, match)
```

## Acknowledgments {.appendix}

Thank you to Prof. Roald Bahr for prioritizing this project. 
Thanks also to the Oslo Sports Trauma Research Centre 
for providing a supportive working environment.

## Author Contributions {.appendix}

Author BC wrote the original code for handling 
the OSTRC-questionnaire and calculating incidences and prevalences. 
Author LKBM generalized BC's code and developed the R package.
