---
title: "Oslo Sports Trauma Research Centre R package"
author:
  - Lena Kristin Bache-Mathiesen^[Oslo Sports Trauma Research Centre, lenakba22@gmail.com]
  - Benjamin Clarsen^[Oslo Sports Trauma Research Centre, benjaminc@nih.no]
output: github_document
---
```{r packages, warning=FALSE, echo = FALSE, message=FALSE}
library(tidyverse) # dplyr, tidyr, ggplot2, purrr and everything else you need to survive
library(ostrc)
```

## Overview
The OSTRC package is a collection of R functions 
for working with sports injury data. Among other 
functions, the package provides tools for handling data from the 
Oslo Sports Trauma Research Centre 
Questionnaire on overuse injuries or health problems. 

### OSTRC questionnaire
The OSTRC questionnaire collects self-reported injury and illness data.   
The functions for handling OSTRC-questionnaire data are 
generalized to work with any OSTRC-questionnaire type and version.
The overuse questionnaire is denoted OSTRC-O. 
The health problem questionnaire is denoted OSTRC-H. 
Version 1.0 corresponds to the original introduction of the questionnaires, 
version 2.0 to the updated version after the review panel meeting.
Based on the articles they were introduced, 
in chronological order from the earliest to latest version:

<ol>
<li>Original overuse questionnaire: 
[OSTRC-O 1.0](doi.org/10.1136/bjsports-2012-091524)</li> 
<li>The expanded version capturing all health problems:
[OSTRC-H 1.0](doi.org/10.1136/bjsports-2012-092087)</li>
<li>The review panel meeting updating both questionnaires:
[OSTRC-O 2.0 and OSTRC-H 2.0](doi.org/10.1136/bjsports-2019-101337)</li>
</ol>

## Installation

```{r install_guide, echo = TRUE, warning = FALSE, eval = FALSE}
# Requires the devtools package
# If devtools is not installed, run code:
# install.packages("devtools") before installing ostrc
devtools::install_github("lenakba/ostrc")
```

## Functions
Below is a brief overview of helpful functions.

### Create case data

The function `create_case_data` finds 
health problems in a dataset with OSTRC-questionnaire responses
and returns a dataframe where one row describes one unique health problem.
The function also finds and adds the startdate, enddate, 
and the duration (in weeks) of each health problem, 
given the date the OSTRC questionnaire was sent.
It also identifies substantial health problems, with the help of 
`find_hp_substantial`, and adds a column for these.

```{r create_case_data_example, echo = TRUE, warning = FALSE}
library(tidyverse) # for tribble() and pipe %>% 
d_ostrc = tribble(~id_participant, ~id_case, ~date_ostrc, ~q1, ~q2, ~q3, ~q4, ~hb_type, ~inj_type,
                  1, 1, "2023-01-01", 8, 0, 17, 25, "Injury", "Overuse",
                  1, 1, "2023-01-07", 8, 0, 17, 25, "Injury", "Overuse",
                  1, 1, "2023-01-14", 8, 0, 17, 0, "Injury", "Overuse",
                  1, 18, "2022-12-07", 25, 0, 0, 0, "Illness", NA,
                  2, 2, "2023-01-12", 8, 8, NA, NA, NA, NA,
                  3, 3, "2022-06-05", 0, 0, 0, 0, NA, NA,
                  4, 4, "2023-01-01", 8, 8, 8, 0, "Injury", "Acute")

# note that the date column must be of class date
d_ostrc = d_ostrc %>% mutate(date_ostrc = as.Date(date_ostrc))

# functions returns one row per health problem
# if a response is not a health problem (like participant ID 3, case ID 3)
# it will not be included in the returned data frame.
d_cases = create_case_data(d_ostrc, id_participant, id_case, date_ostrc, q1, q2, q3, q4)
d_cases
```

Any extra columns in the dataset will be included at the end, 
like columns `hb_type` and `inj_typ` in the example above. Below,
we show what these columns look like after being handled by
the `create_case_data` function.

```{r show_all_cols, echo = TRUE, warning = FALSE}
d_cases %>% select(id_participant, id_case, date_ostrc, hb_type, inj_type)
```

### Find health problems and substantial health problems

Given a vector of OSTRC responses to question 1, 
The function `find_hp` identifies whether or not an observation
is a health problem as per the definition of a health problem
in the OSTRC questionnaire. 

```{r find_hp_example, echo = TRUE, warning = FALSE}
ostrc_1 = c(0, 8, 17, 25)

find_hp(ostrc_1)
```

If the vector of responses to question 1 is not
coded with the standard 0, 8, 17, 25 responses,
but with 1 to 4, or 0 to 3, these can also work.
In these cases, `find_hp` uses the `standardize_codes` 
function automatically.

```{r find_hp_example_codes, echo = TRUE, warning = FALSE}
ostrc_1_other_values = c(0, 1, 2, 3)

find_hp(ostrc_1_other_values)
```

Substantial health problems are more 
complicated to identify, and depends on
the OSTRC version (1.0 or 2.0). 
The function `find_hp_substantial`
identifies substantial injuries given 
question 1, 2, and 3 of the OSTRC questionnaire.
If there is no health problem, it will return as missing.

```{r find_hp_substantial_example, echo = TRUE, warning = FALSE}
ostrc_1 = c(0, 8, 8, 8)
ostrc_2 = c(0, 0, 0, 25)
ostrc_3 = c(0, 0, 17, 0)
   
find_hp_substantial(ostrc_1, ostrc_2, ostrc_3)
```

Make note of how `find_hp_substantial` treats missing in one
or more of the OSTRC-questionnaire items.
If one of the three questions indicate a substantial health problem,
`find_hp_substantial` will consider it substantial, 
regardless of missing values
in other items. No matter whether the missing values are, in reality,
substantial or non-substantial, we still have enough information
to conclude that the observation is substantial.

However, if the only non-missing responses
are values that do not indicate a substantial health problem,
then it will return as missing. In these cases, there is no way of knowing
whether one of the missing observations was, in reality,
a response pertaining to a substantial health problem.

```{r find_hp_substantial_example_missing, echo = TRUE, warning = FALSE}
ostrc_1_missing = c(25, NA, NA, NA)
ostrc_2_missing = c(NA, NA, NA, 17)
ostrc_3_missing = c(NA, 8, NA, NA)

find_hp_substantial(ostrc_1_missing, ostrc_2_missing, ostrc_3_missing)
```


### Find and add event IDs

Function `add_event_id` finds intervals for each event in a longitudinal dataset, and
adds a column indexing them. The intervals may be of unequal lengths. 
This is useful for finding intervals between multiple injuries or
episodes of pain on the same individual, that is, identifying injury cases. 
Or the opposite: the intervals where the participant is currently injured or in pain.

It can also be used to find the
so-called micro-cycle between matches and competition events. 
Micro-cycles are the time from a previous competition or match
to the next competition or match. That is,
all days up to first match/competition in one cycle,
all days from first match to second match in the next cycle and so on. 
Coaches often plan training according to such 
micro-cycles.

This is handy for counting the number of injury cases or number of micro-cycles,
calculating the average interval time, 
or calculating the average of anything else per injury or other event.   

```{r event_id_example, echo = TRUE, warning = FALSE}
library(tidyverse)
library(ostrc)
d_matches_sim = tribble(~date, ~match,
                          "2017-08-22", 0,
                          "2017-08-23", 1,
                          "2017-08-24", 0,
                          "2017-08-25", 0,
                          "2017-08-26", 1)
  d_matches_sim = d_matches_sim %>% mutate(date = lubridate::as_date(date))
  add_event_id(d_matches_sim, date, match)
```

## Acknowledgments {.appendix}

Thank you to Prof. Roald Bahr for prioritizing this project. 
Thanks also to the Oslo Sports Trauma Research Centre 
for providing a supportive working environment.

## Author Contributions {.appendix}

Author BC wrote the original code for handling 
the OSTRC-questionnaire and calculating incidences and prevalences. 
Author LKBM generalized BC's code and developed the R package.
