---
title: "OSTRC package"
author: "Lena Kristin Bache-Mathiesen"
output: github_document
---
```{r packages, warning=FALSE, echo = FALSE, message=FALSE}
library(tidyverse) # life: dplyr, tidyr, ggplot2, purrr and everything else you need to survive
library(ostrc)
```

# Overview
The OSTRC package is a collection of tools for working with sports injury data.

# Installation

```{r install_guide, echo = TRUE, warning = FALSE, eval = FALSE}
# Requires the devtools package
# If devtools is not installed, run code:
# install.packages("devtools") before installing ostrc
devtools::install_github("lenakba/ostrc")
```

# Functions
Below is a brief overview of helpful functions.

## add_micro_cycle
Micro-cycles are the time from a previous competition or match
to the next competition or match. Coaches often plan training according to such 
micro-cycles.
For longitudinal data, `add_micro_cycle` finds the micro-cycles given the date, 
and the event (match or competition).
This is handy for counting the number of micro-cycles
or calculating the average of a variable per micro-cycle, 
among other iterations.   
This can also be used to find each injury interval in the data.

```{r micro_cycle_example, echo = TRUE, warning = FALSE}
d_matches_sim = tribble(~date, ~match,
                          "2017-08-22", 0,
                          "2017-08-23", 1,
                          "2017-08-24", 0,
                          "2017-08-25", 0,
                          "2017-08-26", 1)
  d_matches_sim = d_matches_sim %>% mutate(date = lubridate::as_date(date))
  add_micro_cycle(d_matches_sim, date, match)
```
